Avoiding the Global Sort: \break A Faster Contour Tree Algorithm

### 2017-05-04 paint-contour-tree.tex
figure update

Consider $f: \MM \to \RR$, where $\MM$ is the triangulation of a $d$-dimensional ball with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$. 
The degree assumption can be achieved via vertex unfolding, as discussed later,
%We define this formally later, but it suffices to think of this as a general position assumption.
and is standard in this literature~\cite{kobps-ctsssit-97}.)
% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

We now give a detailed description of the main algorithm.
The primary challenge to use the data structures to rapidly find
ripe vertices.

@@

\section{ \Sec{algo} Algorithm Example }

\textcolor{red}{
In the example below, I assumed that the maxima are already considered ``processed'', so that there are not additional trivial steps.  
This actually makes more sense since the $att()$ should really only ever point to already processed vertex.  Therefore, I suggest 
changing line 3 of init to ``Mark all maxima as processed.  Mark all other critical $v$ as unprocessed.''
Perhaps line 4 needs to be changed also?
}

>>>>

Consider $f: \MM \to \RR$, where $\MM$ is the triangulation of a $d$-dimensional ball with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$, see \Def{deg}. 
The degree assumption can be achieved via vertex unfolding, as discussed later,
%We define this formally later, but it suffices to think of this as a general position assumption.
and is standard in this literature~\cite{kobps-ctsssit-97}.)
% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

We now give a detailed description of the main algorithm.
The primary challenge is to use the data structures to rapidly find
ripe vertices.
An simple example of the execution of the algorithm is given in \Sec{example}.

@@

\section{ \Sec{algo} Algorithm Example }\label{sec:example}

% \textcolor{red}{
% In the example below, I assumed that the maxima are already considered ``processed'', so that there are not additional trivial steps.  
% This actually makes more sense since the $att()$ should really only ever point to already processed vertex.  Therefore, I suggest 
% changing line 3 of init to ``Mark all maxima as processed.  Mark all other critical $v$ as unprocessed.''
% Perhaps line 4 needs to be changed also?
% }

Here we provide a simple example of the execution of the main algorithm from \Sec{algo}.  
An example input mesh is shown in \Fig{mesh1}, followed by the different steps of the join construction in \Fig{mesh2}, 
and finally a table showing the states of the data structures and how they are modified in \Fig{exec}.

Note that in \Fig{exec} a ``stage'' refers to any time at which one of the data structures is modified.  
For simplicity, the trivial stages where the maxima are removed from their corresponding heaps are not shown. 
(Alternatively the algorithm in \Sec{algo} could be modified to exclude the maxima from the heaps, 
but we wanted to keep the algorithm description simple.)
